# **\"Climate spiral\" -- Plotting GISS Surface Temperature**

[NASA](https://www.nasa.gov/) has been collecting surface temperature for more than over 100 years, and the GISS Surface Temperature analysis from 1888 onward. It is an estimate of ***global surface temperature change***.

The temperature analysis schema was defined in the late 1970s by James Hansen and the complete analysis and method are documented in [Hansen and Lebedeff (1987)](https://pubs.giss.nasa.gov/abs/ha00700d.html). To put it simply is a method of global temperature change that is used for comparison with one-dimensional global climate models.

### **Data**

The [website](https://data.giss.nasa.gov/gistemp/) (<https://data.giss.nasa.gov/gistemp/> ) offers loads of data and I am using the ***Global-mean monthly, seasonal, and annual means**, 1880-present, updated through most recent month: [TXT](https://data.giss.nasa.gov/gistemp/tabledata_v4/GLB.Ts+dSST.txt), [CSV](https://data.giss.nasa.gov/gistemp/tabledata_v4/GLB.Ts+dSST.csv)*

It need some prior data preparation, namely min and max values, as it is required by radarchart function. but we need to install these packages first

```{r}
#| eval: false
devtools::install_github(
  repo = "ricardo-bion/ggradar",
  dependencies = TRUE
  )
```


```{r}
library(ggradar)
library(fmsb)
library(scales)
library(RColorBrewer)
```


```{r}
#data txt and preparation
df <-read.csv("data/GLB.Ts+dSST.csv",header = TRUE, sep = ",", skip = 1, dec="." )[1:13]
df <- as.data.frame(sapply(df[1:143,], as.numeric))
df_months <- names(df)[2:13]
df_years <- df$Year
rownames(df) <- df_years
df <- df[,2:13]
```


```{r}
# adding max min
max_min <- data.frame(
  Jan = c(1.4, -0.85), Feb = c(1.4, -0.85), Mar = c(1.4, -0.85),
  Apr = c(1.4, -0.85), May = c(1.4, -0.85), Jun = c(1.4, -0.85),
  Jul = c(1.4, -0.85), Aug = c(1.4, -0.85), Sep = c(1.4, -0.85),
  Oct = c(1.4, -0.85), Nov = c(1.4, -0.85), Dec = c(1.4, -0.85)
)
rownames(max_min) <- c("Max", "Min")
```


```{r}
#merging
df <- rbind(max_min, df)
 
# Set graphic colors
nb.cols <- length(df_years)
mycolors <- colorRampPalette(brewer.pal(8, "Set2"))(nb.cols)
colors_border <- mycolors  
colors_in <- alpha(mycolors, 0.3) 
```


### **Data visualisation**

Adding the radarchart and looping through the years:
```{r}
for (i in 1:length(df_years)){
  y <- df_years[1:i]
  df_tmp <- df[rownames(df)%in%y,1:12]
  df_tmp <- rbind(max_min, df_tmp)
  radarchart( df_tmp, maxmin=TRUE, axistype=1,seg=3,vlabels = df_months,
              plwd=0.5 , plty=1,centerzero=FALSE,caxislabels = c(-1, 0, 1, 1.4),
              cglcol="grey", cglty=2, axislabcol="black",  
              vlcex=1.2,
              title= paste0("GISS Surface temperature for years until ", tail(y,1)) )  
  legend(x=-0.35, y=0.15, legend = tail(y,1), bty = "n", pch=30 , col=colors_in , text.col ="black", cex=1.3, pt.cex=3)
}
```


Changes through time stretch from 1888 until 2023. The scale of this radar charts go from -1ºC to 1.4ºC (with 0 and 1 as reference lines).
